---
title: "Group Project 2 - Exploratory Data Analysis"
author: "Mohamed Jalaly"
date: "3/15/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(kableExtra)
library(scales)
library(skimr)
library(magrittr)
animals = read_csv("dataset11.csv")
```

# Exploratory Data Analysis

First, we would like to understand the distribution of different animal types in our dataset from the shelter. This would give us a better idea if a single type of animal might be influencing the fit of the data.

```{r animaltypes}

animals %>% 
  group_by(animal_type) %>% 
  summarize(counts=n()) %>% 
  mutate(proportion=percent(counts/sum(counts))) %>% 
  kable()

ggplot(animals, mapping=aes(x=animal_type, fill=factor(animal_type))) +  
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=percent)+
  labs(title="Frequency of Animal Type in Shelter Data",
       x="Animal Type", y="Percentage")
  
```

Additionally, we would like to investigate the distribution of animals by the method of arriving at the shelter.

```{r intaketype}

#animals$date = as.yearmon(paste(animals$year, animals$month), "%Y %m")

animals %>% 
  group_by(intake_type) %>% 
  summarize(counts=n()) %>% 
  mutate(proportion=percent(counts/sum(counts))) %>% 
  kable()

ggplot(animals, mapping=aes(x=intake_type, fill=factor(intake_type))) +  
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=percent)+
  labs(title="Frequency of Intake Type in Shelter Data",
       x="Intake Type", y="Percentage")

ggplot(animals, mapping=aes(x=intake_type, fill=factor(animal_type))) +  
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=percent)+
  labs(title="Frequency of Intake Type in Shelter Data",
       x="Intake Type", y="Percentage")
```

Additionally, we want to look at the animals arrival trend to the shelter.

```{r arrivaldistribution, eval=TRUE, echo=TRUE}

# adding a date column
View(animals)
animals %>% 
  group_by(date) %>% 
  summarize(Animals_arriving=n()) %>% 
  ggplot(mapping = aes(x=date, y=Animals_arriving))+
  geom_line(col="blue")+
  labs(title="Animals Arrival Timeseries", x="Date", y="Animal Arrival")

```

Finally, let us look at the summary statistics of the Time at Shelter, which is our variable of interest.

```{r summstat}

animals %>% 
  select(time_at_shelter) %>% 
  skim() %>% 
  kable

animals %>% 
  select(time_at_shelter) %>% 
  ggplot(mapping=aes(x=time_at_shelter))+
  geom_histogram(col="white", binwidth = 3)

###
# split time:
# 0-2: Very Short
# 3-5: short
# 6-8: mid
# 9-11: long
# 11+: Very long
```