---
title: "Group Project 2 - Poisson model"
author: "YueXu Wang & LiJia Wang"
date: "3/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(scales)
library(skimr)
library(ggplot2)
library(tidyverse)
library(lmtest)
library(MASS)
library(pscl)
library(gridExtra)
```

```{r}
animals=read.csv('dataset11.csv')
glimpse(animals)
```

Firstly, we use scatter plot to check if there are some extreme points.
```{r}
p1<-ggplot(animals,mapping=aes(x=intake_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p2<-ggplot(animals,mapping=aes(x=chip_status,y=time_at_shelter))+
  geom_point(col="#f46d43")
p3<-ggplot(animals,mapping=aes(x=animal_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p4<-ggplot(animals,mapping=aes(x=outcome_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p1
p2
p3
p4
```
From the scatter plots, we could see that for variable animal_type, there are several types of animals which are LIVESTOCK, BIRD and WILDLIFE in small numbers.So we want to combine these three types into one called 'others'. We think month and year are meaningless to our analysis. So we will exclude these two variables from our model. 

```{r}
animals$animal_type[which(animals$animal_type=='BIRD')]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="WILDLIFE")]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="LIVESTOCK")]<-"OTHERS"
animals.cleaned=subset(animals,select=-c(month,year))
```

Then we want to check if there is excess zero in our data.
```{r}
zero_set<-animals.cleaned%>%
  filter(time_at_shelter=="0")
nrow(zero_set)
```
We have 380 zeros in our model which is a vary large number so we think there is zero-exceed in our dataset.


# Poisson model
Our task is to find Which factors influence the number of days an animal spends in the shelter before their final outcome is decided. we regard time_at_shelter as response variable and since it is a count data, we intend to use poisson regression model to fit our data. We think month and year are meaningless to our analysis. So we will exclude these two variables from our model.

```{r}
animals.glm.poisson = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned, family=poisson(link="log"))
animals.glm.poisson%>%
  summary()
```

From the output of our model, we could see that outcome_typeFOSTER is not significant, think about the scatter plot of time_at_shelter against outcome_type, there exits a very large time_at_shelter value equal to 69 which maybe an outlier. Build a new datasets without this outlier.

```{r}
animaldata<-animals.cleaned[-617,]
```
```{r}
animals.glm = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animaldata, family=poisson(link="log"))
animals.glm%>%
  summary()
```

We could see all the variables are significant here.


#check if this model has Overdispersion
For poisson model, variance should be equal to mean. We build scatter plots to check this assumption.
```{r}

ggplot(animals.glm, aes(x=log(fitted(animals.glm)), y=log((animaldata$time_at_shelter-fitted(animals.glm))^2)))+ geom_point(col="#f46d43") +geom_abline(slope=1, intercept=0, col="#a6d96a", size=1) + ylab(expression((y-hat(mu))^2)) + xlab(expression(hat(mu)))
```

We could see that a large number of points are above the line which means our model exits overdispersion. There are several methods to deal with overdispersion.

```{r}
animals.glm$deviance/animals.glm$df.residual
```

A guideline is that if the ratio of residual error to residual degrees of freedom exceeds 1.5, the model is overdispersed. So the poisson model is a poor fit.


#Firstly, we use Quasi-Poisson model to deal with overdispersion

```{r}
animals.glm.quasi = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animaldata, family=quasipoisson(link="log"))
animals.glm.quasi%>%
  summary()
drop1(animals.glm.quasi, test = "F")
```

Based on F-test, all variables are significant here and could be included in our model.

```{r}
animals.glm.quasi$deviance/animals.glm.quasi$df.residual
```

Check the ratio again, the quasi-poisson also fit our data not very well.

#Residual plots vs. predicted
```{r}
pred <- predict(animals.glm, type = "response")
stand.resid <- rstandard(model = animals.glm, type = "pearson") # Standardised Pearson residuals
par(mfrow=c(1,2))
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals",
main = "Regular likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")
pred <- predict(animals.glm.quasi, type = "response")
stand.resid <- rstandard(model = animals.glm.quasi, type = "pearson") # Standardised Pearson residuals
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals", main = "Quasi-likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")
```

We can see in the plots that most of the residuals are contained within ±3 in the quasi-Poisson model, while quite a few are outside this range for the original Poisson model.


# neagative binomial model
Another way to solve overdispersion is negative binomial model

```{r}
animals.glm.nb=glm.nb(time_at_shelter~animal_type + intake_type + chip_status+ outcome_type, data=animaldata)
animals.glm.nb%>%
  summary()
```


#compare the poisson model and negative binomial model 

```{r}
#deviance
c(animals.glm$deviance,animals.glm.nb$deviance)
#negative binomial model
c(animals.glm$aic,animals.glm.nb$aic)
#ratio for negative binomial model
animals.glm.nb$devianc/animals.glm.nb$df.residual
```

We can see the deviance and AIC of negative binomial model are much smaller than the poisson one. As for the deviance ratio, we could see that 1.178912 is smaller than 1.5, so negative binomial model fit our data better.


#deal with zero-excess
We intend to use hurdle model. It is a two-part model. The first part is a binary model. This model analyzes whether an observation takes a positive count or not. The second part is zero-truncated models which means we only fit positive counts. Since our positive counts have overdispersion, we choose hurdle negative binomial model to fit our data.
```{r}
animal.hurdle<- hurdle(time_at_shelter~animal_type + intake_type + chip_status+ outcome_type, data=animaldata, dist = "negbin",link = c("logit"))
summary(animal.hurdle)
```


```{r}
sum(predict(animal.hurdle, type = "prob")[,1])
```

we could see that 380 "0" counts hurdle model predict which happens to be the number of zeros in the observed data. 

#check if the hurdle model fit our data well

```{r}
install.packages("countreg", repos="http://R-Forge.R-project.org")
library(countreg)
rootogram(animal.hurdle, max = 80)
```

The line at 0 allows us to easily visualize where the model is over- or under-fitting. At 0 it fits perfectly by design.We could see that counts 3,4 we see little over-fitting (over the line) and then pronounced under-fitting at counts 5 (under the line), but not very much. Check AIC, we could see that aic of hurdle model is lower which means better fit.

```{r}
AIC(animal.hurdle)
```

hurdle model have a smaller AIC value than negative binomial model.

#extra comparsion:Pseudo.R.squared
```{r}
library(rcompanion)
compareGLM(animals.glm,animals.glm.nb)
```
# conclusion
Based on the AIC, we finally use hurdle nagative binomial model to fit our data.

model A:

$$ log(\frac{u_i}{1-u_i})= 7.8792+0.8581 ~ Animal\_type\_DOG-4.7983 ~ Owner\_SURRENDER_1 ~ -3.7933 ~ intake\_type\_STRAY -1.0526 ~ chip\_status\_UNABLE\_TO\_SCAN-2.3051~ outcome\_type\_DIED - 3.5892~outcome\_type\_EUTHANIZED -2.7973 ~ outcome\_typeFOSTER- 4.7999 ~ outcome\_type\_RETURNED\_TO\_OWNER $$

In this model,except for **animal_typeOTHERS** and **chip_statusSCAN NO CHIP**, all others are significant. Let's take animal_type as an example, dog's odds of staying at shelter
is exp(0.8581) times as large as the odds of cat by assuming other variables are held constant. And cat's odds of staying at shelter is exp(0.8236) times as large as the odds of other animals by assuming other variables are held constant.
Another model is:
model B
$$ log(\frac{p_i}{1-p_i}) = 3.18537-1.35700 ~ Owner\_SURRENDER_1~ -0.91826 ~ intake\_type\_STRAY -0.32018 ~ outcome\_type\_EUTHANIZED -1.14834~outcome\_type\_RETURNED\_TO\_OWNER $$
Compared with model A, **animal_typeDOG**,**chip_status_UNABLE TO SCAN**,**outcome_type_FOSTER ** and **outcome_type_DIED** are not significant anymore. It means these variables impact whether or not animals will stay at shelter, but they don't impact how many days they will stay at shelter.
To interpret coefficients,let's take animal_type as an example. The positive count ratio is:
(here put those two equations)
$$\frac{DOGS}{CATS}=exp(0.06531)$$
$$\frac{CATS}{OTHER\_TYPES}=exp(0.31221)$$
Both these two equations are based on the assumption that other variables are held constant.




