---
title: "Group Project 2 - Poisson model"
author: "YueXu Wang & LiJia Wang"
date: "3/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(scales)
library(skimr)
library(ggplot2)
library(tidyverse)
library(lmtest)
library(MASS)
library(pscl)
library(gridExtra)
```

```{r}
animals=read.csv('dataset11.csv')
glimpse(animals)
```

# Poisson model
Our task is to find Which factors influence the number of days an animal spends in the shelter before their final outcome is decided.we regard time_at_shelter as response variable and since it is a count data, we intend to use poisson regression model to fit our data. 

Firstly, we use scatter plot to check if there are some extreme points.
```{r}
p1<-ggplot(animals,mapping=aes(x=intake_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p2<-ggplot(animals,mapping=aes(x=chip_status,y=time_at_shelter))+
  geom_point(col="#f46d43")
p3<-ggplot(animals,mapping=aes(x=animal_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p4<-ggplot(animals,mapping=aes(x=outcome_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p1
p2
p3
p4
```

We also need to check the distribution of our outcome variable time_at_shelter.
```{r}
ggplot(data=animals,aes(x=time_at_shelter))+
  geom_histogram()
```
We could see from the histgram, almost time_at_shelter is between 0 and 10. 
We think month and year are meaningless to our analysis. So we will exclude these two variables. 
From the scatter plots, we could see that for variable animal_type, there are several types of animals which are LIVESTOCK, BIRD and WILDLIFE in small numbers.So we want to combine these three types into one called 'others'. We think month and year are meaningless to our analysis. So we will exclude these two variables from our model. 

```{r}
animals$animal_type[which(animals$animal_type=='BIRD')]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="WILDLIFE")]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="LIVESTOCK")]<-"OTHERS"
animals.cleaned=subset(animals,select=-c(month,year))
```

Then we want to check if there is excess zero in our data.
```{r}
zero_set<-animals.cleaned%>%
  filter(time_at_shelter=="0")
nrow(zero_set)
```
We have 380 zeros in our model which is a vary large number so we think there is zero-exceed in our dataset.

```{r}
animals.glm = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned, family=poisson(link="log"))
animals.glm%>%
  summary()
```
```{r,check goodness-of-fit}
#person statistic
sum(resid(animals.glm ,type="pearson")^2)
#Chi-Square(1667)
qchisq(0.95,1667,ncp=0)
```

Both the deviance 8883.7 and pearson statistic 11632.76 are larger than Chi-Square(1667) 1763.099 which indicate a poor fit.

#check if this model has Overdispersion
For poisson model, variance should equal to mean. We build scatter plots to check this assumption.
```{r}
ggplot(animals.glm, aes(x=log(fitted(animals.glm)), y=log((animals.cleaned$time_at_shelter-fitted(animals.glm))^2)))+ geom_point(col="#f46d43") +geom_abline(slope=1, intercept=0, col="#a6d96a", size=1) + ylab(expression((y-hat(mu))^2)) + xlab(expression(hat(mu)))
```

We could see that a large number of points are above the line which means these exits overdispersion. There are several methods to deal with overdispersion.

Firstly, we use Quasi-Poisson model to deal with overdispersion
#Quasi-Poisson model
```{r}
animals.glm.quasi = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned, family=quasipoisson(link="log"))
animals.glm.quasi%>%
  summary()
drop1(animals.glm.quasi, test = "F")
```
Based on F-test,we intend to drop chip_status first.
```{r}
animals.glm.quasi = glm(time_at_shelter~animal_type + intake_type + outcome_type, data=animals.cleaned, family=quasipoisson(link="log"))
animals.glm.quasi%>%
  summary()
drop1(animals.glm.quasi, test = "F")
```
Now the left variables are all significant.

Compare the deviance of poisson model and quasi-poisson model.
```{r}
c(animals.glm$deviance, animals.glm.quasi$deviance)
```
The deviance of Quasi-Poisson model is bigger than poisson model.

Residual plots vs. predicted
```{r}
pred <- predict(animals.glm, type = "response")
stand.resid <- rstandard(model = animals.glm, type = "pearson") # Standardised Pearson residuals
par(mfrow=c(1,2))
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals",
main = "Regular likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")

pred <- predict(animals.glm.quasi, type = "response")
stand.resid <- rstandard(model = animals.glm.quasi, type = "pearson") # Standardised Pearson residuals
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals", main = "Quasi-likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")
```
We can see in the plots that almost all residuals are contained within ±3 in the quasi-Poisson model, while quite a few are outside this range for the original Poisson model.

Another way to solve overdispersion is negative binomial model
```{r}
animals.glm.nb=glm.nb(time_at_shelter~animal_type + intake_type + chip_status+ outcome_type, data=animals.cleaned)
animals.glm.nb%>%
  summary()
```
compare the poisson model and negative binomial model
```{r}
#deviance 
c(animals.glm$deviance,animals.glm.nb$deviance)
#AIC
c(animals.glm$aic,animals.glm.nb$aic)
```
We can see the deviance and AIC of negative binomial model are much smaller than the poisson one.


#deal with zero-exceed.
```{r}
animal.hurdle<- hurdle(time_at_shelter~animal_type + intake_type + chip_status+ outcome_type, data=animals.cleaned, dist = "negbin",link = c("logit"))
summary(animal.hurdle)
```

```{r}
sum(predict(animal.hurdle, type = "prob")[,1])
```

we could see that 380 "0" counts hurdle model predict which happens to be the number of zeros in the observed data. 

```{r}
#install.packages("countreg", repos="http://R-Forge.R-project.org")
library(countreg)
rootogram(animal.hurdle, max = 80)
```
The line at 0 allows us to easily visualize where the model is over- or under-fitting. At 0 it fits perfectly by design.We could see that counts 3,4 we see dramatic over-fitting (over the line) and then pronounced under-fitting at counts 5 (under the line), but not very much. Check AIC, we could see that aic of hurdle model is lower which means better fit.
```{r}
c(animals.glm.nb$aic,animals.glm.nb$aic)
AIC(animal.hurdle)
```
conclusion: All four variables will influence the time that animals stay in shelter, however, except for cats and dogs, which are common animal types, other animals types have no significant effect on the time they stay in shelter. At the same time, those pets that did not scan the chip tag did not have a significant effect on shelter time.


