---
title: "Group Project 2 - Poisson model"
author: "YueXu Wang"
date: "3/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(scales)
library(skimr)
library(ggplot2)
library(tidyverse)
library(lmtest)
library(MASS)
library(pscl)
library(gridExtra)
```

```{r}
animals=read.csv('dataset11.csv')
glimpse(animals)
```

# Poisson model
Our task is to find Which factors influence the number of days an animal spends in the shelter before their final outcome is decided? So we could regard time_at_shelter as response variable and since it is a count data , we intend to use poisson regression model to fit our data. 


Firstly, we use scatter plot to check if there are some extreme points.
```{r}
p1<-ggplot(animals,mapping=aes(x=intake_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p2<-ggplot(animals,mapping=aes(x=chip_status,y=time_at_shelter))+
  geom_point(col="#f46d43")
p3<-ggplot(animals,mapping=aes(x=animal_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p4<-ggplot(animals,mapping=aes(x=outcome_type,y=time_at_shelter))+
  geom_point(col="#f46d43")
p1
p2
p3
p4
```
From the scatter plots, we could see that for variable animal_type, there are several types of animals which are LIVESTOCK, BIRD and WILDLIFE in small numbers.

We also need to check the distribution of our outcome variable time_at_shelter.

```{r}
ggplot(data=animals,aes(x=time_at_shelter))+
  geom_histogram()
```
We could see from the histgram, almost time_at_shelter is between 0 and 10. 
We think month and year are meaningless to our analysis. So we will exclude these two variables. 
```{r}
animals$animal_type[which(animals$animal_type=='BIRD')]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="WILDLIFE")]<-"OTHERS"
animals$animal_type[which(animals$animal_type=="LIVESTOCK")]<-"OTHERS"
animals.cleaned=subset(animals,select=-c(month,year))

```

Then we want to check if there is excess zero in our data.
```{r}
zero_set<-animals.cleaned%>%
  filter(time_at_shelter=="0")
nrow(zero_set)
```
We have 380 zeros in our model which is a vary large number so we think there is zero-exceed in our dataset.

We think month and year are meaningless to our analysis. So we will exclude these two variables from our model. 

```{r}
animals.glm = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned, family=poisson(link="log"))
animals.glm%>%
  summary()
```
the animal_type and outcome_type variables are not significant in our model.
```{r,check goodness-of-fit}
#person statistic
sum(resid(animals.glm ,type="pearson")^2)
#Chi-Square(1665)
qchisq(0.95,1665,ncp=0)
```

Both the deviance 8872.8 and pearson statistic 11629.56 are larger than Chi-Square(1665) 1761.042 which indicate a poor fit.

#check if this model has Overdispersion
For poisson model, variance should equal to mean. We build scatter plots to check this assumption.
```{r}
ggplot(animals.glm, aes(x=log(fitted(animals.glm)), y=log((animals.cleaned$time_at_shelter-fitted(animals.glm))^2)))+ geom_point(col="#f46d43") +geom_abline(slope=1, intercept=0, col="#a6d96a", size=1) + ylab(expression((y-hat(mu))^2)) + xlab(expression(hat(mu)))
```
```{r,outlier}
animals.outlier=animals.cleaned%>%
  mutate(x=fitted(animals.glm),y=(animals.cleaned$time_at_shelter-fitted(animals.glm))^2)
animals.cleaned2=animals.outlier %>% slice(-c(191,135))
```

```{r}
animals.glm2 = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned2, family=poisson(link="log"))
animals.glm2%>%
  summary()
```

We could see that a large number of points are above the line which means these exits overdispersion. There are several methods to deal with overdispersion.

Firstly, we use Quasi-Poisson model
#Quasi-Poisson model
```{r}
animals.glm.quasi = glm(time_at_shelter~animal_type + intake_type + chip_status+outcome_type, data=animals.cleaned, family=quasipoisson(link="log"))
animals.glm.quasi%>%
  summary()
drop1(animals.glm.quasi, test = "F")
```
Based on F-test,we intend to drop chip_status first.
```{r}
animals.glm.quasi = glm(time_at_shelter~animal_type + intake_type + outcome_type, data=animals.cleaned, family=quasipoisson(link="log"))
animals.glm.quasi%>%
  summary()
drop1(animals.glm.quasi, test = "F")
```
Now the left variables are all significant.
Compare the deviance.
```{r}
c(animals.glm$deviance, animals.glm.quasi$deviance)
```
The deviance of Quasi-Poisson model is smaller than poisson model.

Residual plots vs. predicted
```{r}
pred <- predict(animals.glm, type = "response")
stand.resid <- rstandard(model = animals.glm, type = "pearson") # Standardised Pearson residuals
par(mfrow=c(1,2))
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals",
main = "Regular likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")

pred <- predict(animals.glm.quasi, type = "response")
stand.resid <- rstandard(model = animals.glm.quasi, type = "pearson") # Standardised Pearson residuals
plot(x = pred, y = stand.resid, xlab = "Predicted count", ylab = "Standardised Pearson residuals", main = "Quasi-likelihood", ylim = c(-5,5))
abline(h = c(-3, -2, 0, 2, 3), lty = "dotted", col = "red")
```
We can see in the plots that almost all residuals are contained within ±3 in the quasi-Poisson model, while quite a few are outside this range for the original Poisson model.

```{r}
animals.glm.cleaned$df.null
qchisq(0.95,1655)
```

#2022/3/18

